Declarative Pearl: Lightweight Probabilistic Programming in Curry

### 2015-08-27 Paper/paper.tex
Change type of mkDist
Clean up wet grass example and add debugger tools as contributions

The main goal of probabilistic programming is to open the door for a
wide range of groups that want to evaluate probabilistic models: data
scientists, domain experts and other developers. %
Probabilistic programming languages want to minimise the burden for
non-experts (in the area of machine learning and co) to design their
own model and run algorithm on their data. %
These languages already provide a rich set of primitives and most
importantly inference algorithms that are built upon the results of
experts. %
Thus, a predefined and well-elaborated toolset simplifies the access
to machine learning, because technical details can be ignored. %

Contributions:
\begin{itemize}
\item We present \emph{PFLP} a lightweight library for probabilistic
  programming in the functional logic language Curry. %
  The library's core consists of ... data structures as well as
  ... functions to work on these strucutres, overall the library needs
  less than 30 lines of code. %
\item We can reduce the library code to such a minimum by reusing Curry's
  semantic. %
  The best part is: no further mechanisms are necessary to enable
  probabilistic computations in Curry. %
  That said, we show high connections between current probability
  programming languages and Curry, including lazy evaluation,
  nondeterministic values as well as corresponding built-in search
  abilities and call-time choice. %
\item We present examples that emphasise the benefits of Curry's
  representation of nondeterminsm, which enables early pruning of the
  search space, compared to Haskell's MonadPlus instance for lists,
  which is often used to model nondeterminism. %
\end{itemize}

@@

\begin{figure}
\caption{Bayesian network with conditional probability table for \emph{alarm}-scenario}
\end{figure}

We can model this bayesian network with the following code. %

\begin{verbatim}
earthquake = bernoulli 0.2
burglary   = bernoulli 0.7

alarm True  True  = bernoulli 0.9
alarm True  False = bernoulli 0.8
alarm False True  = bernoulli 0.1
alarm False False = bernoulli 0.0
\end{verbatim}

The definition for the probabilistic functions \code{earthquake},
\code{burglary} and \code{alarm} follow straight-forward from the
scheme of the given model. %
The function \code{bernoulli} yields a probabilistic value that is
\code{True} with probability given by the first argument and
\code{False} with the converse probability. %

\begin{verbatim}
bernoulli prob = mkDist True prob
bernoulli prob = mkDist False (1.0 - prob)
\end{verbatim}

@@

mkDist :: a -> Probability -> Dist a
mkDist = Dist
\end{verbatim}

@@

biasedCoin = (head, 0.3)
biasedCoin = (tail, 0.7)
\end{verbatim}

@@

\begin{verbatim}
fairCoin = (head ? tail,0.5)
\end{verbatim}

Let us now take a look at the bayesian network example again. %
We have already seen the definition of the model in the
introduction. %
However, the more interesting part is to make queries on the given
model that cannot be answered with the probability table itself. %
For example, we can ask the model for the probability that a burglary
occured given the fact that the alarm was set of. %

\begin{verbatim}
-- P(B=T | A=T)
alarmByBurglary =
  burglary >>>= \b ->
  earthquake >>>= \e ->
  alarm b e >>>= \a ->
  guard a >>>= \_ ->
  pure b
\end{verbatim}

@@

\begin{verbatim}
coin' = (True, 0.5) ? (False, 0.5)

coin = (True ? False, 0.5)
\end{verbatim}

>>>>

The main goal of probabilistic programming is to open the door to the
evaluation of probabilistic models for a wide range of users: data
scientists, domain experts and other developers. %
Probabilistic programming languages want to minimise the burden for
non-experts (in the area of machine learning and co) to design their
own model and run inference algorithms on their data. %
These languages already provide a rich set of primitives and most
importantly inference algorithms that are built upon the results of
experts. %
Thus, a predefined and well-elaborated toolset simplifies the access
to machine learning, because technical details can be ignored. %

Contributions:
\begin{itemize}
\item We present \emph{PFLP} a lightweight library for probabilistic
  programming in the functional logic language Curry. %
  The library's core consists of ... data structures as well as
  ... functions to work on these strucutres, overall the library needs
  less than 30 lines of code. %
\item We can reduce the library code to such a minimum by reusing Curry's
  semantic. %
  The best part is: no further mechanisms are necessary to enable
  probabilistic computations in Curry. %
  That said, we show high connections between current probability
  programming languages and Curry, including lazy evaluation,
  nondeterministic values as well as corresponding built-in search
  abilities and call-time choice. %
\item We present examples that emphasise the benefits of Curry's
  representation of nondeterminsm, which enables early pruning of the
  search space, compared to Haskell's MonadPlus instance for lists,
  which is often used to model nondeterminism. %
\item It is fair to assume that a wide range of tools existing for
  Curry can be applied to probabilistic languaeges as well. %
  For instance, \cite{Brassel-BIO} has already presented work on a
  debugger for Curry that works well with nondeterminism and free
  variables. %
\end{itemize}

@@

\begin{figure}
\label{fig:WetGrass}
\caption{Bayesian network with conditional probability table for
  \emph{Wet Grass}-scenario}
\end{figure}

We can model the bayesian network given in Figure \ref{fig:WetGrass} with the following code. %

\begin{verbatim}
rain :: Dist Bool
rain = bernoulli 0.2

sprinkler :: Bool -> Dist Bool
sprinkler False = bernoulli 0.4
sprinkler True  = bernoulli 0.01

grassWet :: Bool -> Bool -> Dist Bool
grassWet False False = bernoulli 0.0
grassWet False True  = bernoulli 0.8
grassWet True  False = bernoulli 0.9
grassWet True  True  = bernoulli 0.99
\end{verbatim}

The definition for the probabilistic functions \code{rain},
\code{sprinkler} and \code{grassWet} follow straight-forward from the
scheme of the given model. %
The function \code{bernoulli} yields a probabilistic value that is
\code{True} with probability given by the first argument and
\code{False} with the converse probability. %

\begin{verbatim}
bernoulli prob = dist True prob
bernoulli prob = dist False (1.0 - prob)
\end{verbatim}

@@

dist :: a -> Float -> Dist a
dist x prob = Dist x (Prob prob)
\end{verbatim}

@@

biasedCoin = dist head 0.3
biasedCoin = disy tail 0.7
\end{verbatim}

@@

\begin{verbatim}
fairCoin = dist (head ? tail) 0.5
\end{verbatim}

Let us now take a look at the bayesian network example again. %
We have already seen the definition of the model in the
introduction. %
However, the more interesting part is to make queries on the given
model that cannot be answered with the probability table itself. %
For example, we can ask the model for the probability that it was
raining when observing wet grass. %

\begin{verbatim}
-- P(B=T | A=T)
grassIsWet =
  rain >>>= \r ->
  sprinkler r >>>= \s ->
  grassWet s r = >>>= \g ->
  guard g >>>= \_ ->
  pure r
\end{verbatim}

\begin{verbatim}
guard :: Bool -> Dist ()
guard True = pure ()
\end{verbatim}

@@

\begin{verbatim}
coin' = mkDist True 0.5 ? mkDist False 0.5

coin = mkDist (True ? False) 0.5
\end{verbatim}
